

```{r}
sum_n_elements <- function(my_vector, n) {
  num_v <- as.numeric(my_vector)
  i <- 0
  chunks <- vector()
  while (i < length(num_v)) {
    chunk_sum <- sum(num_v[(i+1):(i+n)])
    i <- i + n
    chunks <- c(chunks, chunk_sum)
    print(paste(i, chunk_sum))
  }
  return(chunks)
}



#chunks_1000 <- sum_n_elements(akhbarani_first_v, 1000)
chunks_1000 <- colSums(matrix(akhbarani_first_v, nrow=500))
plot(chunks_1000, type="h")
```

























We have seen that the function `which()` can be used to identify the index 
position of a string in a vector. If you click the `run` button under the
code section below, you will see an example of the output of this function:

```{r}
haddathani_v <- which(book_word_v == "حدثني")
haddathani_v
```

The `which()` function returns an integer vector, a sequence of integers (numbers)
that represent the index positions of the strings in the `book_word_v` that match
the condition between the parentheses.
Using the `==` operator means that only positions on which the vector `book_word_v`
contains exactly the string "حدثني" will be added to the `haddathani_v`; if we
want to be a bit more flexible (for example, include tokens in which the word
is preceded by the enclitic conjunction *wa-*), we have to use regular expressions. 

Regular expressions make it possible to define in a more precise way what strings
we want to match (e.g., match only part of a string in a vector, instead of the
entire string). 
A couple of functions can be used in R for regular expressions. Jockers' book 
describes the use of the `grep()` function, but this function does not work 
well for strings that contain characters in non-Latin script. We will use a 
number of functions from the `stringr` package, which is part of the `tidyverse`.

The `str_which()` function takes two arguments: first, the character vector
containing the strings you want to match; and second, the pattern you want to 
match in that string. The function returns an integer vector containing the 
index positions of the strings that matched the pattern in the input vector. 

To see how this works, run the following piece of code: 



(from y value 0 to y value 1) 
for every position in the `match_V` that contains the value `TRUE`.

NB: this kind of plot is called a histogram plot




The `seq()` function creates a sequence of numbers. For example, this code
creates a sequence of 1 to 10.

```{r}
seq(from = 1, to = 10)
```

Modify the code above to create a sequence of numbers, from the first word of 
the book to the last, and assign it to the variable `book_seq_v`: 


```{r ex="seq", type="sample-code"}
# Create a sequence of numbers, from the first word of 
# the book to the last, and assign it to the variable `book_seq_v`


```

```{r ex="seq", type="solution"}
book_seq_v <- seq(from = 1, to = length(book_word_v))
```

```{r ex="seq", type="hint"}
"Use the length() function to get the number of tokens in `book_word_v"`
```

```{r ex="seq", type="sct"}
# does not seem to work...

```

In order to plot the occurrences of a word in the course of the book
(we will use the word *ḥaddathanī/ḥaddathanā*, "he transmitted to us", 
an important term in the context of citation practices in Arabic),
we will first identify which of the positions in this sequence represent the word.

We have seen that the function `which()` can be used to identify the index 
position of a string in a vector. If you click the `run` button under the
code section below, you will see an example of the output of this function:

```{r}
haddathani_v <- which(book_word_v == "حدثني")
haddathani_v
```

The `which()` function returns an integer vector, a sequence of integers (numbers)
that represent the index positions of the strings in the `book_word_v` that match
the condition between the parentheses.
Using the `==` operator means that only positions on which the vector `book_word_v`
contains exactly the string "حدثني" will be added to the `haddathani_v`; if we
want to be a bit more flexible (for example, include tokens in which the word
is preceded by the enclitic conjunction *wa-*), we have to use regular expressions. 

Regular expressions make it possible to define in a more precise way what strings
we want to match (e.g., match only part of a string in a vector, instead of the
entire string). 
A couple of functions can be used in R for regular expressions. Jockers' book 
describes the use of the `grep()` function, but this function does not work 
well for strings that contain characters in non-Latin script. We will use a 
number of functions from the `stringr` package, which is part of the `tidyverse`.

The `str_which()` function takes two arguments: first, the character vector
containing the strings you want to match; and second, the pattern you want to 
match in that string. The function returns an integer vector containing the 
index positions of the strings that matched the pattern in the input vector. 

To see how this works, run the following piece of code: 

```{r}
library("stringr")
test_v <- c("abc", "def", "cbad")
str_which(test_v, "a.")
```

The output of the function shows that the first and third string in the `test_v`
vector contain the character "a" followed by another variable. 

Now, use the `str_which()` function to check which tokens in the `book_word_v` vector
contain either the strings "ḥaddathanā" or "ḥaddathanī". 

NB: you can use the pattern elements `[]` (any of the characters between the brackets)
or `|` (either one string or another). 


```{r ex="str_which", type="pre-exercise-code"}
if(! "stringr" %in% (.packages())) library("stringr")

if !exists(book_word_v) {
  # make sure Arabic is displayed correctly: 
  Sys.setlocale(category = "LC_ALL", locale = "C.UTF-8")
  
  url <- "https://raw.githubusercontent.com/OpenITI/0325AH/master/data/0310Tabari/0310Tabari.Tarikh/0310Tabari.Tarikh.Shamela0009783BK1-ara1.completed"
  text_v <- scan(url, what="character", sep="\n", encoding="UTF-8")
  splitter_index <- which(text_v == "#META#Header#End#")
  lines_v <- text_v[(splitter_index+1):length(text_v)]
  book_v <- paste(lines_v, collapse = "\n")
  book_word_l <- str_split(book_v, "\\W+")
  book_word_v <- unlist(book_word_l)
  
}

```

```{r ex="str_which", type="sample-code"}
# Use the `str_which()` function to check which tokens in the `book_word_v` vector
# contain either the strings "ḥaddathanā" or "ḥaddathanī"
# and assign the output of the function to a variable named `haddathani_v`:


```

```{r ex="str_which", type="solution"}
# Use the `str_which()` function to check which tokens in the `book_word_v` vector
# contain either the strings "ḥaddathanā" or "ḥaddathanī"
# and assign the output of the function to a variable named `haddathani_v`:

haddathani_v <- str_which(book_word_v, "حدثن[ايى]")
```

```{r ex="str_which", type="hint"}
"Use the length() function to get the number of tokens in `book_word_v"`
```

```{r ex="str_which", type="sct"}
# does not seem to work...

```

In order to plot the occurrences of a word in the course of the book
(we will use the word *ḥaddathanī/ḥaddathanā*, "he transmitted to us", 
an important term in the context of citation practices in Arabic),
we will first identify which of the positions in this sequence represent the word.

We have seen that the function `which()` can be used to identify the index 
position of a string in a vector. If you click the `run` button under the
code section below, you will see an example of the output of this function:

```{r}
haddathani_v <- which(book_word_v == "حدثني")
haddathani_v
```















```{r}
# make sure Arabic is displayed correctly: 
Sys.setlocale(category = "LC_ALL", locale = "C.UTF-8")

url <- "https://raw.githubusercontent.com/OpenITI/0325AH/master/data/0310Tabari/0310Tabari.Tarikh/0310Tabari.Tarikh.Shamela0009783BK1-ara1.completed"
t <- scan(url, what="character", sep="\n", encoding="UTF-8")
t[1:50]
```

### Exercise:

In the following script editor (on the left hand side), adapt the script above so that it loads another OpenITI text into a variable called `t_lines`, and print the first 70 lines of that text:

NB: to get the URL for an OpenITI text: 

1. go to the OpenITI corpus search app at [kitab-corpus-metadata.azurewebsites.net/](https://kitab-corpus-metadata.azurewebsites.net/)
2. use the filter field to search for your text
3. find the best version of your text (the one with "PRI" in the `Book Id` column), 
right-click on the book ID in the `Book Id` column, and choose `Copy link address`
in the pop-up menu.

```{r ex="load_text", type="sample-code"}
# make sure Arabic is displayed correctly: 
Sys.setlocale(category = "LC_ALL", locale = "C.UTF-8")

# create a variable `url` and assign the url of your text to it:


# load the text from the url and assign it to a variable called `t_lines`:


# print the first 70 lines of the loaded text:


```

```{r ex="load_text", type="solution"}
# make sure Arabic is displayed correctly: 
Sys.setlocale(category = "LC_ALL", locale = "C.UTF-8")

# create a variable `url` and assign the url of your text to it:
url <- "https://raw.githubusercontent.com/OpenITI/0300AH/master/data/0279Baladhuri/0279Baladhuri.FutuhBuldan/0279Baladhuri.FutuhBuldan.Shamela0012221-ara1.completed"

# load the text from the url and assign it to a variable called `t_lines`:
t_lines <- scan(url, what="character", sep="\n", encoding="UTF-8")

# print the first 70 lines of the loaded text:
t_lines[1:70]

```

```{r ex="load_text", type="sct"}
# does not seem to work...
#test_student_typed("t_lines *<-", not_typed_msg  = "You did not assign a value to variable t_lines")
#test_student_typed("url *<-", not_typed_msg  = "You did not assign a value to variable `url`")
#test_student_typed("url *<- *[\"\']", not_typed_msg  = "You did not use quotes when assigning the link to variable `url`")
#test_output_contains("######OpenITI#", incorrect_msg = "Make sure to print `t_lines`.")

#success_msg("Great!")
```


