---
title: " Exercises on regular expressions, loops and if statements in R"
output: 
  html_document:
      self_contained: false
      css: "style.css"
---

```{r setup, include=FALSE}
tutorial::go_interactive()
```

## Introduction

These exercise will help you practise and develop your newly-acquired skills in 
in R. 

Create a new R script file you name `Chapter4-5_exercise_<yourname>.R` 
And write down the code for your solutions to these exercises. 

Use comments to separate the different exercises in your script. 

## 1. regular expressions : basic exercises

NB: Useful links:

* [Rexegg regular expressions cheat sheet](https://www.rexegg.com/regex-quickstart.html#chars)
* [virtual keyboard](https://pverkind.github.io/arabicVirtualKeyboard/) for 
writing regular expressions containing Arabic script easier (make sure to toggle "RTL regex mode")

### a. Load a new text from OpenITI

Use the [Kitab metadata app](https://kitab-corpus-metadata.azurewebsites.net/) 
to find the best digital text of al-Istakhri's geographical work al-Masalik 
wa-al-Mamalik, and load the text in R. 

NB: 

* for how to use the Kitab metadata app, see [here](https://openiti.github.io/2021IslamicateWorldCourse/week-6-interacting-with-the-openiti-corpus-using-github-and-kate.html#finding-texts-in-the-openiti-corpus).
* To get the link to a text, right-click the Version Id in the right-most column
and choose `copy link address`.

```{r ex="setting_up", type="pre-exercise-code"}
# make sure Arabic is displayed correctly: 
Sys.setlocale(category = "LC_ALL", locale = "C.UTF-8")
```

```{r ex="setting_up", type="sample-code"}
# Load the text using its URL and remove the metadata header: 

```

```{r ex="setting_up", type="solution"}
library("readr")
# Load the text using its URL and remove the metadata header: 
url <- "https://raw.githubusercontent.com/OpenITI/0350AH/master/data/0346Istakhri/0346Istakhri.MasalikWaMamalik/0346Istakhri.MasalikWaMamalik.Shamela0011680-ara1.mARkdown"
text_lines_v <- read_lines(url)
splitter_pos <- which(text_lines_v == "#META#Header#End#")
body_lines_v <- text_lines_v[(splitter_pos + 1):length(text_lines_v)]
```


### b. regular expressions exercise 1

Use a regular expression to find all the lines in the text that contain the word
"درهم" (dirham, "silver coin") or its plural "دراهم" (darāhim)

How many times is the word used? 

Visualize with a dispersion plot where in the text it is used. 


### c. regular expressions exercise 2

Use a regular expression to find all the lines in the text that contain the word
"إلى", "to(wards)" (note that the last letter of ilā changes into a yā' ("ي")
if it is followed by a suffix; we want to include those cases as well)



## Loops

### basic loop exercise: 

Create a vector that contains at least 4 numbers, and print for every 
number in that vector its index position in the vector, and its value. 

```{r}

```


### basic loop exercise 2:

Create a vector `source_v` that contains at least 4 character strings. 
Create a loop that prints the number of characters in each character string 
and stores those counts in a new vector called `character_counts_v`.

You can use the `nchar()` function to count the number of characters in a string. 
(Ask shaykh Google or the R help function how to use `nchar()`)


```{r}

```


### basic loop exercise 3: 

Create a vector `source_v` that contains at least 4 character strings. 
Use a "for loop" and an "if...else condition" to store the character counts
of strings shorter than 5 characters in a vector `short_v`, and the others in
a vector called `long_v`. 

```{r}

```

## More advanced regular expressions exercise 3: use the `str_extract_all()` function to find words in context


The `str_extract_all()` function creates a character vector that contains
all matches for a search pattern in a given character vector. 

```{r}
library(stringr)
ex <- c("I love apples", "I like pears", "I hate bananas")

# find all words in the vector that contain the letter "e"
result_l <- str_extract_all(ex, "\\w*e\\w*")
str(result_l)
# turn the list of results into a character vector:
result_v <- unlist(result_l)
str(result_v)
```

What does the output of the `str()` function tell you about the variables `result_l` 
and `result_v`?

Use this same technique to extract all mentions of the word "فارس" from the text. 

```{r}

```

This is not very informative: you get only a sequence of repetitions of the word Fārs. 

Would it not be great if we could print the context of the word as well? Say, 
20 characters before and 20 characters after the search word? 

Write a regular expression that will match the word Fārs + 20 characters before
it and 20 characters after it. Then use a loop to print each of the search results
in its context. 

NB: do not use the vector that contains the text split into lines for this, 
but first collapse the lines into one character string!

```{r}

```



## More advanced loop exercise: 

OpenITI texts contain milestone markers after every 300 Arabic tokens; they look like this: 
`ms001` (after the first 300 Arabic tokens), `ms002` (after the second 300 Arabic tokens),
etc.

Use Jockers' method from chapter 5 to divide the text not into chapters but 
into 300-word chunks, and plot how often the words "منبر" (pulpit, for mosque) 
and مدينة (city) are mentioned in each chunk. 


```{r}

```

### More looping: repeating the same code for multiple files: 

Ask shaykh google how to save a plot created by the plot() function as an image 
file. When you found the solution, create dispersion plots for a word in 4 text files 
of your choice using a loop, and save each of those plots to a file.

```{r}

```